---
title: "Data Cleaning- Glassdoor Dataset"
author: "Shefali C."
date: "2023-10-27"
output: 
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This notebook contains steps on data manipulation in R. Most of the steps involve handling text data as the dataset used comprises mostly of textual columns.  
The dataset can be downloaded from [here on kaggle](https://www.kaggle.com/datasets/rashikrahmanpritom/data-science-job-posting-on-glassdoor).

```{r load-libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(janitor)
```

```{r read-data}
#working directory
working_dir <- getwd()

#read the unclean csv file
data <- readr::read_csv(file.path(working_dir, 
                                  "data/DS_Job_Postings_Glassdoor/Uncleaned_DS_jobs.csv"))

```

## Structure of dataframe

**`str(data)`** returns a list of attributes about the dataframe. These include class, column names, data type of the columns and problems with the df.  

```{r data-struc}
str(data)
```

Using **`attributes(data)`** returns a similar list with:  
  - row.names  
  - names (name of columns)  
  - spec (specifications about col-type)  
  - problems, and  
  - class.  


```{r data-attr1}
attributes(data)
```

Each of these individual attributes can also be accessed like this:  
**`attr(data, "spec")`** for only specifications.  
**`attr(data, "class")`** for class of the object and so on...  
  
```{r data-attr2}
attr(data, "class")
attr(data, "spec")
```
  

These attributes give similar result using individual functions like:  
  - **`class(data)`**  
  - **`spec(data)`**  
  - **`names(data)`**  
  
